---
title: "EDA"
author: "Alyssa Vanderbeek"
date: "12/6/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

cancer = read_csv('../data/Cancer_Registry.csv') %>%
  janitor::clean_names() %>%
  select(avg_ann_count:binned_inc)

cancer %>% skimr::skim()

```

I looked at the following variables:

TARGET_deathRate: Dependent variable. Mean per capita (100,000) cancer mortalities(a)

avgAnnCount: Mean number of reported cases of cancer diagnosed annually(a)

avgDeathsPerYear: Mean number of reported mortalities due to cancer(a)

incidenceRate: Mean per capita (100,000) cancer diagnoses(a)

medianIncome: Median income per county (b)

popEst2015: Population of county (b)

povertyPercent: Percent of population in poverty (b)

studyPerCap: Per capita number of cancer-related clinical trials per county (a)

binnedInc: Median income per capita binned by decile (b)


In examining the associations between variables, we see that the outcome of interest ("target_death_rate") is linearly correlated with a county's cancer incidence rate, median income, and percentage of the population in poverty. 

However, we can see that there's some association between covariates. Avg_ann_count, avg_deaths_per_year, and pop_est2015 are nearly perfectly correlated. Med_income and poverty_percent are also related, as can be seen by an exponential relationship. 

```{r}
cancer %>% select(-binned_inc) %>% pairs
```


After looking at the distribution of the variables here, we should perform a log transformation on avg_ann_count, avg_deaths_per_year, and pop_est2015. The other variables are relatively normally distributed. 

```{r}
cancer = cancer %>%
  mutate(log_ann_count = log(avg_ann_count),
         log_deaths_yr = log(avg_deaths_per_year),
         log_pop = log(pop_est2015))
```

